An interdisciplinary examination of stress and injury occurrence in athletes

### 2020-07-20 apa_style.tex
pre results edits

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This is just an example/guide for you to refer to when submitting manuscripts to Frontiers, it is not mandatory to use Frontiers .cls files nor frontiers.tex  %
% This will only generate the Manuscript, the final article will be typeset by Frontiers after acceptance.
%                                              %
%                                                                                                                                                         %
% When submitting your files, remember to upload this *tex file, the pdf generated with it, the *bib file (if bibliography is not within the *tex) and all the figures.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Version 3.4 Generated 2018/06/15 %%%
%%% You will need to have the following packages installed: datetime, fmtcount, etoolbox, fcprefix, which are normally inlcuded in WinEdt. %%%
%%% In http://www.ctan.org/ you can find the packages and how to install them, if necessary. %%%

\documentclass[utf8]{frontiersHLTH}

%\setcitestyle{square} % for Physics and Applied Mathematics and Statistics articles
\usepackage{url,hyperref,lineno,microtype,subcaption}
\usepackage[onehalfspacing]{setspace}

\linenumbers

% BELOW TAKEN FROM rticles plos template
%
% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

% hyperref package, useful for hyperlinks
\usepackage{hyperref}

% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

% Sweave(-like)
\usepackage{fancyvrb}
\DefineVerbatimEnvironment{Sinput}{Verbatim}{fontshape=sl}
\DefineVerbatimEnvironment{Soutput}{Verbatim}{}
\DefineVerbatimEnvironment{Scode}{Verbatim}{fontshape=sl}
\newenvironment{Schunk}{}{}
\DefineVerbatimEnvironment{Code}{Verbatim}{}
\DefineVerbatimEnvironment{CodeInput}{Verbatim}{fontshape=sl}
\DefineVerbatimEnvironment{CodeOutput}{Verbatim}{}
\newenvironment{CodeChunk}{}{}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

\usepackage{color}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Below is from frontiers
%
\bibliographystyle{frontiersinSCNS}
% Use doublespacing - comment out for single spacing
%\usepackage{setspace}
%\doublespacing

% Leave a blank line between paragraphs instead of using \\

\def\keyFont{\fontsize{8}{11}\helveticabold }

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW

%% END MACROS SECTION

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\def\Authors{
  Dr Harry Fisher\,\textsuperscript{1*},
  Dr Marianne Gittoes\,\textsuperscript{1},
  Professor Lynne Evans\,\textsuperscript{1},
  Miss Leah Bitchell\,\textsuperscript{1},
  Dr Richard Mullen\,\textsuperscript{2},
  Dr Marco Scutari\,\textsuperscript{3}}

\def\Address{

  \textsuperscript{1} School of Sport and Health Science, Cardiff Metropolitan University,  Cardiff,   United Kingdom

  \textsuperscript{2} Division of Sport, Health \& Exercise Sciences, Brunel University,  London,   United Kingdom

  \textsuperscript{3} Istituto Dalle Molle di Studi sull'Intelligenza Artificiale (IDSIA),  Manno,   Switzerland
  }

  \def\corrAuthor{Dr Harry Fisher}\def\corrAddress{Cardiff Metropolitan University\\Cyncoed Road\\, CF23 6XD United Kingdom}\def\corrEmail{\href{mailto:harryfisher21@gmail.com}{\nolinkurl{harryfisher21@gmail.com}}}
  \def\firstAuthorLast{Fisher {et~al.}}

\begin{document}
\onecolumn
\firstpage{1}

\title[Stress and sports injury]{An interdisciplinary examination of stress and injury occurrence in
athletes}
\author[\firstAuthorLast]{\Authors}
\address{} %This field will be automatically populated
\correspondance{} %This field will be automatically populated

\extraAuth{}% If there are more than 1 corresponding author, comment this line and uncomment the next one.
%\extraAuth{corresponding Author2 \\ Laboratory X2, Institute X2, Department X2, Organization X2, Street X2, City X2 , State XX2 (only USA, Canada and Australia), Zip Code2, X2 Country X2, email2@uni2.edu}

\maketitle

\begin{abstract}

This paper adopts a novel, interdisciplinary approach to explore the relationship between psychosocial factors, physiological stress-related markers and occurrence of injury in athletes using a repeated measures design across a 2-year data collection period. At three data collection time-points, athletes completed measures of major life events, the reinforcement sensitivity theory personality questionnaire, muscle stiffness, heart rate variability and postural stability, an
d reported any injuries they had sustained since the last data collection. Two Bayesian networks were used to examine the relationships between variables and model the changes between data collection points in the study. Findings revealed muscle stiffness to have the strongest relationship with injury occurrence, with high levels of stiffness increasing the probability of sustaining an injury. Negative life events did not increase the probability of injury occurrence at any single time-point; however, when examining changes between time points, increases in negative life events did increase the probability of injury. In addition, the combination of increases in negative life events and muscle stiffness resulted in the greatest probability of sustaining an injury. Findings demonstrated the importance of both an interdisciplinary approach and a repeated measures design to furthering our understanding of the relationship between stress-related  markers and injury occurrence.

\tiny
 \keyFont{ \section{Keywords:} Sports Injury, Stress, Interdisciplinary, Bayesian Network}

\end{abstract}

\hypertarget{introduction}{%
\section*{Introduction}\label{introduction}}
\addcontentsline{toc}{section}{Introduction}

Over the last four decades sport related injuries have received
increased research attention (Ivarsson et al., 2017). This attention is
unsurprising given the high incidence (Rosa et al., 2014; Sheu et al.,
2016), and undesirable physical and psychological effects of sports
injuries (Leddy et al., 1994; Brewer, 2012). To mitigate against both
the increasing incidence and undesirable consequences of injury,
research has identified several psychological (Slimani et al., 2018),
anatomical (Murphy et al., 2003), biomechanical (Neely, 1998; Hughes,
2014) and environmental (Meeuwisse et al., 2007) factors associated with
sports injury occurrence. Indeed, several models of injury causation
have been proposed that highlight the multifactorial nature of injury
occurrence (Kumar, 2001; Meeuwisse et al., 2007; Wiese-Bjornstal, 2009),
of which one of the most widely cited was developed by Williams and
Anderson (Fig \ref{fig:fig1}; Andersen and Williams, 1988; Williams and
Andersen, 1998).

Williams and Andersen's (Williams and Andersen, 1998) model proposed
that when faced with a potentially stressful athletic situation, an
athlete's personality traits (e.g., hardiness, locus of control and
competitive trait anxiety), history of stressors (e.g., major life
events and previous injuries) and coping resources (e.g., general coping
behaviours) will contribute to their response, either interactively or
in isolation. Central to the model is the stress response, which
reflects the bi-directional relationship between athletes' appraisal of,
and response to, a stressful athletic situation. The model predicts that
athletes who have a history of many stressors, personality traits that
intensify the stress response and few coping resources, will exhibit
greater attentional (e.g., peripheral narrowing) and/or physiological
(e.g., increased muscle tension) responses that put these individuals at
greater risk of injury.

Within Williams and Andersen's (Williams and Andersen, 1998) model,
major life events, a component of an athlete's history of stressors,
most consistently predicts injury occurrence (Williams and Andersen,
2007); specifically, major life events with a negative, as opposed to
positive, valence (Passer and Seese, 1983; Maddison and Prapavessis,
2005). However, personality traits and coping resources have also been
found to predict injury, with for example, athletes more likely to
sustain an injury if they have poor social support and psychological
coping skills, and high trait anxiety and elevated competitive state
anxiety; compared to athletes with the opposite profile. (Smith et al.,
1990; Lavallée and Flint, 1996; Ivarsson and Johnson, 2010). However,
the amount of variance explained by the psychosocial factors proposed by
the model has been modest, typically between 5 - 30\% (Galambos et al.,
2005; Ivarsson and Johnson, 2010); suggesting other factors are also
likely to contribute to injury occurrence.

While the psychosocial factors proposed in Williams and Andersen's
(Williams and Andersen, 1998) model have received the most research
attention, the mechanisms through which these factors are proposed to
exert their effect have remained under-investigated in the literature.
To elaborate, the model suggests that injuries are likely to occur
through either increased physiological arousal resulting in increased
muscle tension and reduced flexibility or attentional deficits caused by
increased distractibility and peripheral narrowing. However, to date,
the research has largely focused on attentional deficits (Andersen and
Williams, 1999; Rogers and Landers, 2005; Wilkerson, 2012; Swanik et
al., 2007). For example, Andersen and Williams (Andersen and Williams,
1999) measured peripheral and central vision during high and low stress
conditions and found athletes with high life event stress coupled with
low social support had greater peripheral narrowing under stressful
conditions compared to athletes with the opposing profile; these
athletes went on to sustain an increased number of injuries during the
following season. Indeed, Rodgers and Landers (Rogers and Landers, 2005)
supported Andersen and Williams's (Andersen and Williams, 1999) earlier
findings reporting that peripheral narrowing under stress mediated 8.1\%
of the relationship between negative life events and injury. However,
the remaining variance between negative life events and athletic injury
through the other proposed mechanisms, such as increased muscle tension
and reduced motor control, remains to be explored (cf.~Williams and
Andersen, 1998).

One challenge faced by researchers addressing the sports injury problem
with a psychological lens is the multifactorial nature of injury, and
the possible contribution of other non-psychological factors to the
stress response (Meeuwisse et al., 2007; Wiese-Bjornstal, 2009). For
example, a large body of research indicates that training-related stress
is also likely to be related to the stress response and injury
occurrence (Lee et al., 2017; Djaoui et al., 2017), and may account for
the unexplained variance from the psychological predictors of injury.
Appaneal and Perna (Appaneal and Perna, 2014) proposed the
biopsychosocial model of stress athletic injury and health (BMSAIH) to
serve as an extension to Williams and Andersen's (Williams and Andersen,
1998) model and to address some of these issues. To elaborate, the
BMSAIH aimed to clarify the mediating pathways between the stress
response and injury, consider other health outcomes and behavioural
factors that impact sports participation, and integrate the impact of
training on athletes' health (Appaneal and Perna, 2014). The central
tenet of the BMSAIH is that psychosocial distress (e.g., negative life
events) may act synergistically with training-related stress as a result
of high-intensity and high-volume sports training, and ``widen the
window of susceptibility'' (Appaneal and Perna, 2014, 74) to a range of
undesirable health outcomes including illness and injury. Consequently,
the BMSAIH provides a framework for future research to build on Williams
and Andersen's (Williams and Andersen, 1998) model, by including other
physiological markers of training-related stress, which together may
provide greater insight into the injury process.

Although research supporting the BMSAIH has mainly focused on the
relationship between hormonal responses to training and injury
occurrence (Perna and McDowell, 1995; Perna et al., 1997, 2003), other
research has identified additional markers of training-related stress
that are associated with an increased risk of injury; for example, heart
rate variability (Bellenger et al., 2016; Williams et al., 2017),
postural stability (Romero-Franco et al., 2014) and muscle stiffness
(Pruyn et al., 2015). However, these markers are often studied in
isolation without an assessment of the psychosocial factors that are
known to contribute to injury, thereby limiting our understanding of how
psychosocially and physiologically derived stress may contribute
synergistically to injury occurrence. Recently, Bittencourt et
al.~(Bittencourt et al., 2016) suggested that to better understand the
multifactorial nature of sports injuries, research needs to move away
from studying risk factors in isolation and instead adopt a complex
systems approach to injury. Such an approach posits that injury may
arise from a complex ``web of determinants'' (Bittencourt et al., 2016,
3), where different factors interact in unpredictable and unplanned
ways, but result in a global outcome pattern of either adaptation or
injury.

A challenge when adopting a complex systems approach is using an
appropriate analysis technique that is able to capture the uncertainty
and complexity of the relationships between different variables. One
technique that provides a solution to this challenge is Bayesian network
(BN) modelling. BN's allow the construction of graphical probabilistic
models using the underlying structure that connects different variables
(nodes in the network) (Scutari and Denis, 2014). The learned structure
can be used for inference by obtaining the posterior probabilities of a
particular node for a given query (e.g., if the value of Node A is x and
the value of Node B is y, what is the probability Node C being value
z?). Furthermore, BN's do not distinguish between dependent and
independent variables as they are a form of unsupervised learning, which
is a strength over regression or structure equation models when the
underlying relationship in the network may not be known (Olmedilla et
al., 2018).

To summarise, despite offering a possible framework to build on the
research stemming from Williams and Andersen's (Williams and Andersen,
1998) model, there remains and opportunity to explore other
physiological stress-related markers proposed by the BMSAIH, in addition
to the already well-established psychological characteristics known to
be related to injury (Appaneal and Perna, 2014). Furthermore, research
has typically not captured changes in both psychosocial factors and
stress-related physiological markers that may occur between initial
measurement and injury occurrence. Given the exploratory nature of the
current study the following objectives were defined:

\begin{itemize}
\tightlist
\item
  Identify suitable markers of stress that can be easily captured in a
  large cohort of athletes in a field based setting.
\item
  Capture the markers of stress and injury occurrence using a
  prospective, repeated measures design.
\item
  Explore and evaluate the relationships between the markers of stress
  and injury using Bayesian network modelling.
\end{itemize}

\hypertarget{methods}{%
\section*{Methods}\label{methods}}
\addcontentsline{toc}{section}{Methods}

\hypertarget{participants}{%
\subsection*{Participants}\label{participants}}
\addcontentsline{toc}{subsection}{Participants}

A total of 351 athletes (male: \emph{n} = 231, female: \emph{n} = 120)
were initially recruited for the study. Participants had an average age
of 22.0 \(\pm\) 7.0 years and represented a range of team (football,
rugby, netball, cricket, lacrosse, basketball and field hockey) and
individual sports (athletics, tennis, weightlifting, gymnastics, judo,
swimming and golf from a British University and local sports clubs
(Table \ref{tab:table1}). Participants self-rated competitive level
ranged from recreational to international standard. A total of 126
(49.03\%) participants had sustained an injury in the 12 months prior to
the start of the study (male: \emph{n} = 114 {[}49\%{]}, female;
\emph{n} = 48 {[}40\%{]}). At the start of the study, all participants
were injury free (no modifications to their usual training routine due
to a sport related medical problem for a minimum of four weeks).
Participants were engaged in training for their respective sports for at
least five hours per week. Ethical approval was obtained from the
University ethics committee prior to the start of the study and all
participants provided informed consent.

@@

\hypertarget{measures}{%
\subsection*{Measures}\label{measures}}
\addcontentsline{toc}{subsection}{Measures}

\hypertarget{major-life-events}{%
\subsubsection*{Major life events}\label{major-life-events}}
\addcontentsline{toc}{subsubsection}{Major life events}

A modified version of the Life Events Survey for Collegiate Athletes
(LESCA) was used to measure participants' history of life event stress
(Petrie, 1992). Modifications were made to the LESCA to ensure the
suitability of the items for the study population (S1 Table). The LESCA
comprises 69 items that reflect possible life events that participants
may have experienced. Example items include, ``Major change in the
frequency (increased or decreased) of social activities due to
participation in sport'', ``Major change in the amount (more or less) of
academic activity (homework, class time, etc)'' and ``Major change in
level of athletic performance in actual competition (better or worse)''.
Participants were asked to rate the perceived impact of each life event
they had experienced within the last 12 months on an 8-point Likert
scale anchored at -4 (\(extremely\;negative\)) and +4
(\(extremely\;positive\)). Negative and positive life event scores were
calculated by summing the negative and positive scores respectively. A
score for total life events was calculated by summing the absolute
values for both negative and positive events. Petrie (1992) (Petrie,
1992) reported test-retest reliabilities at 1-week and 8-weeks with
values ranging from .76 to .84 (p \textless{} .001) and .48 to .72 (p
\textless{} .001) respectively. Petrie also provided evidence of
discriminant, convergent and predictive validity. The LESCA is the most
widely used measure of major life events for athletes in the sports
injury literature. For this study, Composite Reliability (Fornell and
Larcker, 1981) was preferred to Cronbach's alpha as it does not assume
parallelity (i.e., all factor loadings are constrained to be equal, and
all error variances are constrained to be equal) and instead takes into
consideration the varying factor loadings of the items in the
questionnaire. The composite reliability for the LESCA in this study was
0.84.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={An interdisciplinary examination of stress and injury occurrence in athletes},
  pdflang={en-EN},
  pdfkeywords={Sports Injury, Stress, Interdisciplinary, Bayesian Network},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

% Table formatting
\usepackage{longtable}
\usepackage{lscape}
% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
\usepackage{multirow}		% Table styling
\usepackage{tabularx}		% Control Column width
\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
\usepackage{threeparttablex}            % Lets threeparttable work with longtable

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\begin{center}\begin{threeparttable}}
%   {\end{threeparttable}\end{center}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}

% Enables adjusting longtable caption width to table width
% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

% \setlength{\parindent}{0.5in}
% \setlength{\parskip}{0pt plus 0pt minus 0pt}

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\makeatother
\shorttitle{Stress and sports injury}
\author{Harry Fisher\textsuperscript{1}, Dr Marianne Gittoes\textsuperscript{1}, Professor Lynne Evans\textsuperscript{1}, Miss Leah Bitchell\textsuperscript{1}, Dr Richard Mullen\textsuperscript{2}, \& Dr Marco Scutari\textsuperscript{3}}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} Cardiff Metropolitan University, Cardiff, United Kingdom\\\textsuperscript{2} Brunel University, London, United Kingdom\\\textsuperscript{3} Istituto Dalle Molle di Studi sull'Intelligenza Artificiale (IDSIA), Manno, Switzerland}
\authornote{

Correspondence concerning this article should be addressed to Harry Fisher, Cyncoed Road, Cardiff, CF23 6XD. E-mail: harryfisher21@gmail.com}
\keywords{Sports Injury, Stress, Interdisciplinary, Bayesian Network}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifxetex
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[shorthands=off,main=english]{babel}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {\setlength{\parindent}{0pt}%
  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
  {\par}

\title{An interdisciplinary examination of stress and injury occurrence in athletes}

\date{}

\abstract{
This paper adopts a novel, interdisciplinary approach to explore the relationship between psychosocial factors, physiological stress-related markers and occurrence of injury in athletes using a repeated measures design across a 2-year data collection period. At three data collection time-points, athletes completed measures of major life events, the reinforcement sensitivity theory personality questionnaire, muscle stiffness, heart rate variability and postural stability, and reported any injuries they had sustained since the last data collection. Two Bayesian networks were used to examine the relationships between variables and model the changes between data collection points in the study. Findings revealed muscle stiffness to have the strongest relationship with injury occurrence, with high levels of stiffness increasing the probability of sustaining an injury. Negative life events did not increase the probability of injury occurrence at any single time-point; however, when examining changes between time points, increases in negative life events did increase the probability of injury. In addition, the combination of increases in negative life events and muscle stiffness resulted in the greatest probability of sustaining an injury. Findings demonstrated the importance of both an interdisciplinary approach and a repeated measures design to furthering our understanding of the relationship between stress-related markers and injury occurrence.
}

\begin{document}
\maketitle

\section{Introduction}

Over the last four decades sport-related injuries have received increased research attention (Ivarsson et al., 2017) in response to the high incidence (Rosa et al., 2014; Sheu et al., 2016) and associated undesirable physical and psychological effects (Leddy et al., 1994; Brewer, 2012). Multiple psychological (Slimani et al., 2018), anatomical (Murphy et al., 2003), biomechanical (Neely, 1998; Hughes, 2014) and environmental (Meeuwisse et al., 2007) factors have been associated with sports injury occurrence and several models of injury causation have been proposed that highlight the multifactorial nature of injury occurrence (Kumar, 2001; Meeuwisse et al., 2007; Wiese-Bjornstal, 2009), of which one of the most widely cited was developed by Williams and Anderson (Fig \ref{fig:fig1}; Andersen and Williams, 1988; Williams and Andersen, 1998).

Williams and Andersen's (Williams and Andersen, 1998) stress-injury model proposed that when faced with a potentially stressful athletic situation, an athlete's personality traits (e.g., hardiness, locus of control and competitive trait anxiety), history of stressors (e.g., major life events and previous injuries) and coping resources (e.g., general coping behaviours) contribute to the injury response, either interactively or in isolation. The stress response is central to the model and reflects the bi-directional relationship between athletes' appraisal of, and response to, a stressful athletic situation. The model predicts that athletes who have a history of stressors, personality traits that intensify the stress response and few coping resources, will exhibit greater attentional (e.g., peripheral narrowing) and/or physiological (e.g., increased muscle tension) responses that put these individuals at greater injury risk.

Within Williams and Andersen's (Williams and Andersen, 1998) model, major life events, which is a component of an athlete's history of stressors, most consistently predicts injury occurrence (Williams and Andersen, 2007); specifically, major life events with a negative, as opposed to positive, valence (Passer and Seese, 1983; Maddison and Prapavessis, 2005). Personality traits and coping resources have also been found to predict injury with athletes more likely to sustain an injury if they have poor social support and psychological coping skills, and high trait anxiety and elevated competitive state anxiety; compared to those with the opposing profile (Smith et al., 1990; Lavallée and Flint, 1996; Ivarsson and Johnson, 2010).However, the amount of variance explained by these psychosocial factors has been modest and typically between 5 - 30\% (Galambos et al., 2005; Ivarsson and Johnson, 2010), which indicates a likely interaction with other factors.

While the psychosocial factors in Williams and Andersen's (Williams and Andersen, 1998) model have received the most research attention, less insight into the mechanisms through which these factors are proposed to exert their effect exists. To elaborate, the model suggests that injuries are likely to occur through either increased physiological arousal resulting in increased muscle tension and reduced flexibility or attentional deficits caused by increased distractibility and peripheral narrowing. However, to date, the research has largely focused on attentional deficits (Andersen and Williams, 1999; Rogers and Landers, 2005; Wilkerson, 2012; Swanik et al., 2007). For example, Andersen and Williams (Andersen and Williams, 1999) found athletes with high life event stress coupled with low social support had greater peripheral narrowing under stressful conditions compared to athletes with the opposing profile; these athletes went on to sustain an increased number of injuries during the following athletic season. Rodgers and Landers (Rogers and Landers, 2005) further supported Andersen and Williams's (Andersen and Williams, 1999) earlier findings by identifying that peripheral narrowing under stress mediated 8.1\% of the relationship between negative life events and injury.

Knowledge of the physiological factors (e.g., increased muscle tension and reduced motor control) contributing to the remaining variance between negative life events and athletic injury remains sparse (cf.~Williams and Andersen, 1998). One challenge faced by researchers addressing the sports injury problem through a psychological lens is the multifactorial nature of injury, and the possible interaction with physiological factors in the stress response (Meeuwisse et al., 2007; Wiese-Bjornstal, 2009). For example, a large body of research has suggested that training-related stress is also likely to contribute to the stress response and injury occurrence (Lee et al., 2017; Djaoui et al., 2017) and may account for the unexplained variance from the psychological predictors. Appaneal and Perna (Appaneal and Perna, 2014) proposed the biopsychosocial model of stress athletic injury and health (BMSAIH) to serve as an extension to Williams and Andersen's (Williams and Andersen, 1998) model. The BMSAIH enhances our understanding of the mediating pathways between the stress response and injury alongside other health outcomes and behavioral factors that impact sports participation (Appaneal and Perna, 2014). The central tenet of the BMSAIH is that psychosocial distress (e.g., negative life events) may act synergistically with training-related stress as a result of high-intensity and high-volume sports training, and ``widen the window of susceptibility'' (Appaneal and Perna, 2014, 74) to a range of undesirable health outcomes including illness and injury. Consequently, the BMSAIH provides an important framework that has enhanced insight into the multi-faceted nature of the injury process by building on Williams and Andersen's (Williams and Andersen, 1998) model whilst including other physiological markers of training-related stress.

Although research supporting the BMSAIH has mainly focused on the relationship between hormonal responses to training and injury occurrence (Perna and McDowell, 1995; Perna et al., 1997, 2003), other training-related stress markers including heart rate variability (Bellenger et al., 2016; Williams et al., 2017), muscle stiffness (Pruyn et al., 2015) and postural stability (Romero-Franco et al., 2014) have been reported to be associated with an increased risk of injury. However, the largely mono-disciplinary approach to examining the role of these markers has provided limited insight into the multifaceted interactions with psychological factors that may contribute to injury occurrence. Furthermore, a reliance on designs that capture a single point of measurement precludes the assessment of intra as well as inter-individual changes and the effect of the time interval between measurement and injury occurrence on subsequent injuries (Johnson et al., 2014). Such an approach fails to capture changes in both psychological factors and physiological markers that may occur preceding an injury. Importantly, a repeated measurement approach would enable an assessing of how variables and their interactions change with respect to time provide greater insight into the effect that repeated exposure to major life events and other stress-related factors has on injury occurrence.

Recently, Bittencourt et al.~(Bittencourt et al., 2016) advocated a move away from studying isolated risk factors and instead, adopt a complex systems approach in order to understand injury occurrence. Such an approach posits that injury may arise from a complex ``web of determinants'' (Bittencourt et al., 2016, 3), where different factors interact in unpredictable and unplanned ways, but result in a global outcome pattern of either adaptation or injury. Capturing the uncertainty and complexity of the relationships between different variables using an appropriate interdisciplinary analysis within the framework of a complex systems approach is challenging. Bayesian network (BN) modelling provides one solution by allowing the construction of graphical probabilistic models using the underlying structure that connects different variables (Scutari and Denis, 2014). The learned BN structure can be used for inference by obtaining the posterior probabilities of a particular variable for a given query (e.g., if the value of variable A is x and the value of variable B is y, what is the probability variable C of being value z?). Furthermore, unlike regression or structural equation models, BN's do not distinguish between dependent and independent variables when the underlying relationship in the network may not be known (Olmedilla et al., 2018). BN modelling subsequently provides a valuable but underused interdisciplinary approach to investigating the complex and unpredictable interactions of psychological and physiological factors implicated in the injury process.

Using the frameworks provided by Williams and Andersen's (Williams and Andersen, 1998) stress injury model and Appaneal's BMSAIH model, the aim of this interdisciplinary study was to develop new understanding of the multifaceted interactions of psychological and physiological stress-related factors with injury occurrence. A prospective, repeated measures design incorporating field-based physiological and psychological measures combined with a BN modelling analysis was used to address the study aim.

\section{Methods}

\subsection{Participants}

A total of 351 athletes (male: \emph{n} = 231, female: \emph{n} = 120) were initially recruited for the study from a British University and local sports clubs (Table \ref{tab:table1}). Participants had an average age of 22.0 \(\pm\) 7.0 years and represented a range of team (football, rugby, netball, cricket, lacrosse, basketball and field hockey) and individual sports (athletics, tennis, weightlifting, gymnastics, judo, swimming and golf.
Participants self-rated competitive level ranged from recreational to international standard.
They were engaged in training for their respective sports for at least five hours per week.
A total of 126 (49.03\%) participants had sustained an injury in the 12 months prior to the start of the study (male: \emph{n} = 114 {[}49\%{]}, female; \emph{n} = 48 {[}40\%{]}).
All participants were injury free (no modifications to their usual training routine due to a sport related medical problem for a minimum of four weeks) at the start of the study.
Ethical approval was obtained from the University ethics committee prior to the start of the study and all participants provided informed consent.

@@

\hypertarget{measures}{%
\subsection{Measures}\label{measures}}

\hypertarget{injury}{%
\subsubsection{Injury}\label{injury}}

\hypertarget{major-life-events}{%
\subsubsection{Major life events}\label{major-life-events}}

A modified version of the Life Events Survey for Collegiate Athletes (LESCA) was used to measure participants' history of life event stress (Petrie, 1992).
The LESCA is the most widely used measure of major life events for athletes in the sports injury literature.
Modifications were made to the LESCA to ensure the suitability of the items for the study cohort (S1 Table).
The LESCA comprises 69 items that reflect possible life events that participants may have experienced. Example items include, ``Major change in the frequency (increased or decreased) of social activities due to participation in sport'', ``Major change in the amount (more or less) of academic activity (homework, class time, etc)'' and ``Major change in level of athletic performance in actual competition (better or worse)''. Participants were asked to rate the perceived impact of each life event they had experienced within 12-months preceding the study onset on an 8-point Likert scale anchored at -4 (\(extremely\;negative\)) and +4 (\(extremely\;positive\)).
Negative and positive life event scores were calculated by summing the negative and positive scores, respectively. A score for total life events was also calculated by summing the absolute values for both negative and positive events.
Petrie (1992) (Petrie, 1992) reported test-retest reliabilities at 1-week and 8-weeks with values ranging from .76 to .84 (p \textless{} .001) and .48 to .72 (p \textless{} .001) respectively. Petrie also provided evidence of discriminant, convergent and predictive validity. For this study, composite reliability (Fornell and Larcker, 1981) was preferred to Cronbach's alpha as it does not assume parallelity (i.e., all factor loadings are constrained to be equal, and all error variances are constrained to be equal) and instead takes into consideration the varying factor loadings of the items in the questionnaire. The composite reliability for the LESCA in this study was .84.

\hypertarget{reinforcement-sensitivity-theory-personality-questionnaire}{%
\subsubsection{Reinforcement Sensitivity Theory Personality Questionnaire}\label{reinforcement-sensitivity-theory-personality-questionnaire}}
